Effectful Semantics for Substructural SSA

### 2023-11-02 pldi24/paper.tex
Defined substructural monoidal category

\TODO{define a substructural effectful category, rewrite as ``substructural effectful category \(\mc{C}\)" equipped with proper structrure}

>>>>

\begin{definition}[Substructural Monoidal Category]
  A \textbf{substructural monoidal category} is a tuple \((\mc{C}, \mc{A}, \mc{R})\) where:
  \begin{itemize}
    \item \(\mc{C}\) is a family of categories \(\mc{C}^q\) parametrized by subsets \(q \subseteq \{\ms{a}, \ms{r}\}\) such that
    \begin{itemize}
      \item For all \(q \subseteq q'\), \(\mc{C}^{q'}\) is a wide subcategory of \(\mc{C}^q\), and in particular all object sets \(|\mc{C}^q|\) are the same. We will write this set as \(|\mc{C}|\)
      \item All \(\mc{C}^q\) share symmetric monoidal structure, and are individually closed under the tensor products of morphisms
      \item \(\mc{C}^{q \cap q'} = \mc{C}^q \cap \mc{C}^{q'}\)
    \end{itemize}
    \item \(\mc{A} \subseteq |\mc{C}|\) is a set of objects closed under tensor products, such that
    \begin{itemize}
      \item \(\mb{1} \in \mc{A}\)
      \item Every \(A\) in \(\mc{A}\) is equipped with a morphism \(\ms{drop}(A): \mc{C}^{\{\ms{a}, \ms{r}\}}(A, \mb{1})\)
      \item \(\forall A, B \in \mc{A}, \ms{drop}(A \otimes B) = \ms{drop}(A) \otimes \ms{drop}(B);\lambda_I\)
      \item \(\forall A, B \in \mc{A}, \forall f: \mc{C}^{\{\ms{a}\}}(A, B), f;\ms{drop}(B) = \ms{drop}(A)\)
    \end{itemize}
    \item \(\mc{R} \subseteq |\mc{C}|\) is a set of objects closed under tensor products, such that
    \begin{itemize}
      \item \(\mb{1} \in \mc{R}\)
      \item Every \(A\) in \(\mc{R}\) is equipped with a morphism \(\ms{split}(A): \mc{C}^{\{\ms{a}, \ms{r}\}}(A, A \otimes A)\)
      \item \(\forall A, B \in \mc{R}, \ms{split}(A \otimes B) = \ms{split}(A) \otimes \ms{split}(B);\alpha;A \otimes \sigma_{A, B} \otimes B;\alpha\)
      \item Associativity: \(\forall A \in \mc{R}, \ms{split}(A);\ms{split}(A) \otimes A;\alpha = \ms{split}(A);A \otimes \ms{split}(A)\)
      \item Commutativity: \(\forall A \in \mc{R}, \ms{split}(A);\sigma = \ms{split}(A)\)
      \item Comonoid: \(\forall A \in \mc{A} \cap \mc{R}, \ms{split}(A);A \otimes \ms{drop}(A) = \rho_A^{-1} \quad \ms{split}(A);\ms{drop}(A) \otimes A = \lambda_A^{-1}\)
    \end{itemize}
  \end{itemize}
  We call a substructural monoidal category a \textbf{substructural Cartesian category} if \(\mc{A} = \mc{R} = |\mc{C}|\), in which case \(\mc{C}^{\{\ms{a}, \ms{r}\}}\) is Cartesian. Many of our examples will satisfy this structure.
\end{definition}
A simple example of a substructural Cartesian category is the category of relations \(\ms{Rel}\), viewed as the Kleisli category of the power-set monad with morphisms of the form \(\ms{Rel}(A, B) = A \to \mc{P}(B)\). In particular, we can define:
\begin{equation}
  \begin{aligned}
    \ms{Rel}^\varnothing(A, B) 
    &= \ms{Rel}(A, B) 
    &&= A \to \mc{P}(B) \\
    \ms{Rel}^{\{\ms{a}\}}(A, B) 
    &= \{R \in \ms{Rel}(A, B) \mid \forall a, |R(a)| \geq 1\} 
    &&= A \to \mc{P}^+(B) \\
    \ms{Rel}^{\{\ms{r}\}}(A, B) 
    &= \{R \in \ms{Rel}(A, B) \mid \forall a, |R(a)| \leq 1\} 
    &&= A \to \ms{Option}(B) \\
    \implies \ms{Rel}^{\{\ms{a}, \ms{r}\}}(A, B) 
    &= \{R \in \ms{Rel}(A, B) \mid \forall a, |R(a)| = 1\} 
    &&= A \to B
  \end{aligned}
\end{equation}
Here, \(\ms{split}(A)\) is just the copy operator \(a \mapsto (a, a)\) and \(\ms{drop}(A)\) the deletion operator \(a \mapsto ()\); it is quite trivial to prove that these lie in \(\ms{Rel}^{\{\ms{a}, \ms{r}\}}\), being functions, and that they satisfy the axioms of a commutative comonoid, as desired.

Note that we cannot simply have \(\ms{Rel}^{\{\ms{a}\}} = \ms{Rel}\), since, for any morphism \(R: \ms{Rel}(A, B)\) such that \(\exists a \in A, R(a) = \varnothing\), we have
\begin{equation}
  (R;\ms{drop}(B))(a) = \varnothing \neq \ms{drop}(A)(a) = \{()\}
\end{equation}
We will see that this corresponds to the fact that the following programs are not semantically equal
\begin{equation}
  \ms{let}\;x = \ms{loop}(); e \not\simeq e \qquad \text{where} \qquad x \notin \ms{fv}(e)
\end{equation}
Similarly, we cannot simply have \(\ms{Rel}^{\{\ms{r}\}} = \ms{Rel}\), since, for any morphism \(R: \ms{Rel}(A, B)\) such that \(\exists a \in A, |R(a)| \supseteq \{b_0, b_1\}\) where \(b_0 \neq b_1\), we have that
\begin{equation}
  (R;\ms{split}(B))(a) = \{(b, b) \mid b \in R(a)\} \neq (\ms{split}(A);R)(a) = R(a) \times R(a) 
\end{equation}
since only the left-hand side contains the pair \((b_0, b_1)\), this corresponding to the fact that the following programs are not semantically equal
\begin{equation}
  \ms{let}\;x = \ms{rand}(); (x, x) \not\simeq (\ms{rand}(), \ms{rand}())
\end{equation}
Rather than directly give a concrete example of a general substructural monoidal category, it makes more sense to first jump to the next definition, which will provide us with one more naturally:
\begin{definition}[Susbtructural Effectful Category]
  \ \TODO{this}
\end{definition}

\TODO{state monad}

\TODO{separation logic}
